Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils

### 2020-10-11 doc/Supplementary.Rmd
accepted version

---
title: "Supporting information for Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
output: 
  bookdown::word_document2:
    reference_docx: ijctemplate.docx
bibliography: refs.bib
biblio-style: international-journal-of-climatology.csl
link-citations: yes
always_allow_html: true
---

@@

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse); library(flextable); library(lubridate)
load("../results/logs.RData")

@@

```{r table1, echo = FALSE}
regulartable(ttests) %>% width(width = 1) %>% set_caption("Table S1: Results of the t-tests comparing the paired waterlogged-dry soil bioclimatic variables. Analysis performed with full-year, contemporaneous data series (2015-2017).")
```

```{r table3, echo = FALSE}
regulartable(chelsa) %>% width(width = 1) %>% set_caption("Table S2: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy). Analysis performed with full-year, contemporaneous data series (2015-2017).")
```

>>>>

---
title: "Supporting information for Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
output: github_document
bibliography: refs.bib
biblio-style: international-journal-of-climatology.csl
link-citations: yes
always_allow_html: true
---

@@

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse); library(lubridate)
load("../results/logs.RData")

@@

```{r table1, echo = FALSE}
ttests %>% kableExtra::kable(caption = "**Table S1** *Results of the t-tests comparing the paired waterlogged-dry soil bioclimatic variables. Analysis performed with full-year, contemporaneous data series (2015-2017).*", digits = 3, format = "pandoc")
```

```{r table3, echo = FALSE}
chelsa %>% kableExtra::kable(caption = "**Table S2** *Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy). Analysis performed with full-year, contemporaneous data series (2015-2017).*", format = "pandoc")
```

### 2020-10-11 doc/manuscript.Rmd
accepted version

---
title: "Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
subtitle: "Short title: Testing the thermal buffer of mire groundwater"
output: 
  bookdown::word_document2:
    reference_docx: ijctemplate.docx
bibliography: refs.bib
biblio-style: international-journal-of-climatology.csl
link-citations: yes
always_allow_html: true
---

@@

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse); library(flextable)
load("../results/logs.RData")
load("../results/MSoutput.RData")
```

@@

# Financial support

@@

This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (Fig. \@ref(fig:fig1)a). Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds. In the study area, average annual precipitation ranges from 800 mm at the low elevations to 1800 mm at the high mountains, while average annual temperature ranges from 12.5 ºC to 5 ºC (source = Agencia Estatal de Meteorología, http://www.aemet.es/es/serviciosclimaticos/). Under the most extreme emissions scenario, the latest projections expect temperatures to increase by 4 ºC and precipitation to decrease by c. 10 % by the end of the century, although the precipitation projections are subjected to great uncertainty [@RN4939]. The present humid and mild climate of this region harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246]. Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes. Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition. Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens (https://eunis.eea.europa.eu/habitats/279) on acid bedrocks, to alkaline calcareous fens (https://eunis.eea.europa.eu/habitats/277) on limestone.

We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (Table 1). Although we classify two of these sites as rain-fed bogs based on their vegetation (*Oxycocco-Sphagnetea* Br.-Bl. et Tx. ex Westhoff et al. 1946), we must note that these bogs are very poorly developed on hummocks over valley mires and are in tight connection to the underlying water table coming from streams. Therefore, soil waterlogging in all our sites is fed by streams or springs. In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil. In this ecosystem, it is at this depth that true soil begins to develop, under the porous upper layers made up of live mosses. We installed one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (Fig. \@ref(fig:fig1)b). The vegetation at the sampling points was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points. Dataloggers recorded temperature once every hour and stayed on site for approximately five years, after which we retrieved them and downloaded their records. At the moment of retrieval, the internal clock of all dataloggers had not deviated for more than four hours.

@@

To test if the differences between dry and waterlogged points were significant, we used paired t-tests (n = 8 paired sites per each of the four bioclimatic variables). Tests were one-tailed, according to the original hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range. To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition). We calculated RMSE using its implementation in the package *Metrics* [@RN4941]. We checked whether our data complied with the normality assumption of the t-test and Pearson's correlation using the Shapiro-Wilk normality test; the assumption was met in all cases (p > 0.05). To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary Tables S1 and S2).

@@

Dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one. Time series showed considerably less variation at waterlogged points, for all eight site comparisons (Fig. \@ref(fig:fig2)).

Bioclimatic variables (Table 2 and Fig. \@ref(fig:fig3)) supported this notion. The mean annual range was wider at dry points, as was the mean diurnal range. Maximum temperatures of the warmest months were usually higher at dry measuring points. The opposite was true for the minimums of the coldest months, in which case temperature was generally colder at dry points, although the difference was less pronounced than for the maximums (Fig. \@ref(fig:fig3)).

Patterns regarding the maximum temperatures and annual range were especially noticeable at both low (El Molinucu, La Malva) and high (La Recoleta) elevations. The pattern for the diurnal range, however, was more prominent just at the low sites (El Molinucu, La Malva), whereas the minimums of the coldest months showed no specific pattern (Table 2).

@@

The CHELSA climatic model (Table 3 and Fig. \@ref(fig:fig4)) provided air temperatures that significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84). The correlation between CHELSA and soil values was marginally significant (p = 0.05) in the case of the maximums of the warmest month, but only at dry points (Pearson's r = 0.71). The values of CHELSA did not correlate with soil temperatures in the rest of the cases (p > 0.05). RMSE indicated that the predictions of CHELSA were more accurate at dry points for the cases of the annual range, the diurnal range and the winter min; the CHELSA values were more accurate at waterlogged points in the case of the summer max.

@@

In these southern mires, the buffering effect is much stronger during the summer (Fig. \@ref(fig:fig5)). This is most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva shows especially high summer temperatures within the dry soils (> 40 ºC). It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939]. Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.). This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e. a former glacial lake undergoing silting. However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer. More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.

@@

@@

# Tables

```{r table1, echo = FALSE}
regulartable(table1) %>% width(width = 1) %>% set_caption("Table 1: Mire sites included in this study, indicating the type of fen, pH, conductivity, elevation, coordinates, and length of the temperature recording period. Groundwater pH and conductivity values are the average of 5–10 measures per site taken with a MM40 Portable Multimeter (Crison Instruments S.A., Alella, Spain). Coordinates are in decimal degrees WGS84.")
```

```{r table2, echo = FALSE}
regulartable(table2) %>% width(width = 1) %>% set_caption("Table 2: Groundwater buffer effect per mire and bioclimatic variable. The buffer effect was calculated as the difference between the value in the waterlogged and the dry points.")
```

```{r table3, echo = FALSE}
regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Comparison of soil bioclimatic variables with CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).")
```

# Figures

```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "(a) Location of the study sites in northern Spain. (b) Representation of the sampling design at El Riotuertu site. The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.", fig.width=7, fig.height=3.5}
knitr::include_graphics("../results/Fig1.png")
```

```{r fig2, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Hourly soil temperature records at the mire sites. The blue series was recorded within the mire, in a waterlogged area. The red series was recorded in a neighbouring dry area. Dataloggers were buried at 5 cm depth. The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.", fig.width=7, fig.height=10}
knitr::include_graphics("../results/Fig2.png")
```

```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Average bioclimatic variables in the dry and waterlogged points. The bars represent the mean value, and the brackets the standard error of 8 records.", fig.width=7, fig.height=3}
knitr::include_graphics("../results/Fig3.png")
```

```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs. soil temperatures measured in situ, in dry and waterlogged points.", fig.width=7, fig.height=7}
knitr::include_graphics("../results/Fig4.png")
```

```{r fig5, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "Complementary temporal analysis of the groundwater buffer effect during peak hot events. Variation of soil temperatures at paired sites (dry vs. waterlogged points) during exceptionally warm days of the warm season. Exceptionally warm days were considered those with a daily maximum (at the dry point) belonging to the 90th percentile of daily maximums for the site. The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e. soil temperatures at the waterlogged point are cooler than the neighbouring dry site).", fig.width=3.5, fig.height=3.5}
knitr::include_graphics("../results/Fig5.png")
```

>>>>

---
title: "Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils"
subtitle: "Short title: Testing the thermal buffer of mire groundwater"
output: github_document
bibliography: refs.bib
biblio-style: international-journal-of-climatology.csl
link-citations: yes
always_allow_html: true
---

@@

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse)
load("../results/logs.RData")
load("../results/MSoutput.RData")
```

@@

@@

This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (**Fig. 1a**). Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds. In the study area, average annual precipitation ranges from 800 mm at the low elevations to 1800 mm at the high mountains, while average annual temperature ranges from 12.5 ºC to 5 ºC (source = Agencia Estatal de Meteorología, http://www.aemet.es/es/serviciosclimaticos/). Under the most extreme emissions scenario, the latest projections expect temperatures to increase by 4 ºC and precipitation to decrease by c. 10 % by the end of the century, although the precipitation projections are subjected to great uncertainty [@RN4939]. The present humid and mild climate of this region harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246]. Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes. Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition. Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens (https://eunis.eea.europa.eu/habitats/279) on acid bedrocks, to alkaline calcareous fens (https://eunis.eea.europa.eu/habitats/277) on limestone.

```{r fig1, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 1** **(a)** *Location of the study sites in northern Spain.* **(b)** *Representation of the sampling design at El Riotuertu site. The position of the dry and waterlogged temperature loggers is indicated, the blue area represents the extent of the waterlogged mire habitat.*", fig.width=7, fig.height=3.5}
knitr::include_graphics("../results/Fig1.png")
```

We selected `r length(unique(temperatures$Logger))/2` mire sites representing the regional elevation gradient of mire vegetation and the different mire types (**Table 1**). Although we classify two of these sites as rain-fed bogs based on their vegetation (*Oxycocco-Sphagnetea* Br.-Bl. et Tx. ex Westhoff et al. 1946), we must note that these bogs are very poorly developed on hummocks over valley mires and are in tight connection to the underlying water table coming from streams. Therefore, soil waterlogging in all our sites is fed by streams or springs. In each site, we buried two dataloggers (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil. In this ecosystem, it is at this depth that true soil begins to develop, under the porous upper layers made up of live mosses. We installed one datalogger in a flat waterlogged spot within the mire; the other one in the close vicinity, but in a flat and dry upland area outside the mire (**Fig. 1b**). The vegetation at the sampling points was always either mire or pasture, with no shrubs, trees or any other landscape features shading the measuring points. Dataloggers recorded temperature once every hour and stayed on site for approximately five years, after which we retrieved them and downloaded their records. At the moment of retrieval, the internal clock of all dataloggers had not deviated for more than four hours

```{r table1, echo = FALSE}
table1 %>% kableExtra::kable(caption = "**Table 1** *Mire sites included in this study, indicating the type of fen, pH, conductivity, elevation, coordinates, and length of the temperature recording period. Groundwater pH and conductivity values are the average of 5–10 measures per site taken with a MM40 Portable Multimeter (Crison Instruments S.A., Alella, Spain). Coordinates are in decimal degrees WGS84.*", format = "pandoc")
```

@@

To test if the differences between dry and waterlogged points were significant, we used paired t-tests (n = 8 paired sites per each of the four bioclimatic variables). Tests were one-tailed, according to the original hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range. To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition). We calculated RMSE using its implementation in the package *Metrics* [@RN4941]. We checked whether our data complied with the normality assumption of the t-test and Pearson's correlation using the Shapiro-Wilk normality test; the assumption was met in all cases (p > 0.05). To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary **Tables S1** and **S2**).

@@

Dataloggers recorded temperatures for five years in five of the sites, four years in two, and two years in one. Time series showed considerably less variation at waterlogged points, for all eight site comparisons (**Fig. 2**).

```{r fig2, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 2** *Hourly soil temperature records at the mire sites. The blue series was recorded within the mire, in a waterlogged area. The red series was recorded in a neighbouring dry area. Dataloggers were buried at 5 cm depth. The amplitude of diurnal temperature range in dry sites was reduced with increasing elevation, where the thermal buffer effect was also less apparent.*", fig.width=7, fig.height=10}
knitr::include_graphics("../results/Fig2.png")
```

Bioclimatic variables (**Table 2** and **Fig. 3**) supported this notion. The mean annual range was wider at dry points, as was the mean diurnal range. Maximum temperatures of the warmest months were usually higher at dry measuring points. The opposite was true for the minimums of the coldest months, in which case temperature was generally colder at dry points, although the difference was less pronounced than for the maximums (**Fig. 3**).

```{r fig3, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 3** *Average bioclimatic variables in the dry and waterlogged points. The bars represent the mean value, and the brackets the standard error of 8 records.*", fig.width=7, fig.height=3}
knitr::include_graphics("../results/Fig3.png")
```

Patterns regarding the maximum temperatures and annual range were especially noticeable at both low (El Molinucu, La Malva) and high (La Recoleta) elevations. The pattern for the diurnal range, however, was more prominent just at the low sites (El Molinucu, La Malva), whereas the minimums of the coldest months showed no specific pattern (**Table 2**).

@@

```{r table2, echo = FALSE}
table2 %>% kableExtra::kable(caption = "**Table 2** *Groundwater buffer effect per mire and bioclimatic variable. The buffer effect was calculated as the difference between the value in the waterlogged and the dry points.*", format = "pandoc")
```

The CHELSA climatic model (**Table 3** and **Fig. 4**) provided air temperatures that significantly correlated (p < 0.05) with soil temperatures only in the case of the minimums of the coldest month, both at dry points (Pearson's r = 0.90) and waterlogged points (Pearson's r  = 0.84). The correlation between CHELSA and soil values was marginally significant (p = 0.05) in the case of the maximums of the warmest month, but only at dry points (Pearson's r = 0.71). The values of CHELSA did not correlate with soil temperatures in the rest of the cases (p > 0.05). RMSE indicated that the predictions of CHELSA were more accurate at dry points for the cases of the annual range, the diurnal range and the winter min; the CHELSA values were more accurate at waterlogged points in the case of the summer max.

```{r table3, echo = FALSE}
table3 %>% kableExtra::kable(caption = "**Table 3** *Comparison of soil bioclimatic variables with CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).*", format = "pandoc")
```

```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 4** *Scatter plots of the bioclimatic variables predicted by CHELSA air temperatures vs. soil temperatures measured in situ, in dry and waterlogged points.*", fig.width=7, fig.height=7}
knitr::include_graphics("../results/Fig4.png")
```

@@

In these southern mires, the buffering effect is much stronger during the summer (**Fig. 5**). This is most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva shows especially high summer temperatures within the dry soils (> 40 ºC). It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939]. Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.). This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e. a former glacial lake undergoing silting. However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer. More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.

```{r fig5, echo = FALSE, message = FALSE, warning = FALSE, fig.pos = "H", fig.cap = "**Figure 5** *Complementary temporal analysis of the groundwater buffer effect during peak hot events. Variation of soil temperatures at paired sites (dry vs. waterlogged points) during exceptionally warm days of the warm season. Exceptionally warm days were considered those with a daily maximum (at the dry point) belonging to the 90th percentile of daily maximums for the site. The 1:1 line (corresponding to no buffer effect) is shown, points falling to the right of this line show a buffer effect (i.e. soil temperatures at the waterlogged point are cooler than the neighbouring dry site).*", fig.width=3.5, fig.height=3.5}
knitr::include_graphics("../results/Fig5.png")
```

@@

# Acknowledgements

@@
